---
title: "Plot Superstar"
author: "AMS"
date: '2022-07-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data visualization:

-   The practice of translating information into a visual context

-   Helps make complex data easier to understand

### Goals:

-   Understand the different types of variables

-   To plot or not?

-   Produce useful and meaningful graphics in R

-   Understand when and when not to use a graphic.

-   Lots of coding

Bonus:

-   Plotting neuroimaging data in R

-   Natural language processing in R

-   

    -   

        ```{}
        ```

### Libraries

The required libraries for this tutorial include `ggplot2` . As learned in previous lectures, please install and load the library below.

Good practice when working in rmarkdowns is to `#comment` out any installation lines of code, because some installs can be lengthy.

```{r}
#install ggplot2 below.
replace_me_with_code
#load the library ggplot2 below
replace_me_with_code
```

```{r}
#install ggplot2 below.
#install.packages('ggplot2')
#load the library ggplot2 below
library(ggplot2)
library(tidyverse)
```

### Load data

```{r}
df = read.csv("https://raw.githubusercontent.com/rbcavanaugh/pitt-csd-r-workshop/master/DataViz/osf_meir%20and%20Bangash.csv")
```

### Variables:       

-   Categorical: A characteristic of someone of something that can be called by name (e.g., handedness)

    -   Types of plots for categorical variables:

        Pie plot:

        ```{}
        ```

```{}
```

To provide some new code, today we will load data directly from github. Instead of providing a path on your local machine, we will just provide the website and the type of datafile.

#### About the data

The data to be loaded below comes from 34 individuals with aphasia.

```{}
```

### Take a peak at the data 

There are a couple ways to look at your data structure once it has been loaded. we will use the summary function below. This dataset includes information from 34 people with aphasia and includes 33 different variables. As you will see when you run the summary function, the output is not always easy to parse when there are a large number of variables.

#### Run the summary function by inserting the dataframe name into the the summary function

```{r}
summary()
```

Solution

```{r}
summary(df)
```

#### Here is a screen grab from [osf](https://osf.io/3eg4x/wiki/home/)

#### Now that we have a loaded the dataset let's jump into plotting. Whenever a plot is generated, it should convey a message that is not easily done so by words. 

### Visualize 

```{}
```

### What did you learn? Did we need it?

### 

#### The first question you should think about when starting the data visualization processing is "why." Why do you want to look at your data. 

### Consider the third variable in the data set, `handedness`. 
