---
title: "Exploratory data analysis"
author: "AMS"
date: '2022-07-01'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(dplyr)
```

### Where we were?

-   Installed R and R studio

-   Learned to install packages/load packages

-   Learned about R environments (super helpful, not regularly taught. Be bold!)

-   Learned how to load data sets

-   Name data sets

### To the future, exploratory data analysis:
> Exploratory data analysis is the first step in most research projects. Generally, exploratory data analysis, or EDA,  is a task that statisticians, data scientists, and researchers complete to visualize their data graphically or tabula to help:
1. Gain a basic understanding of the central tendencies of your variables
2. Explore the relationships between variable
3. Generate and answer questions about your data

> The ultimate goal of EDA is to use what you have learned to refine research questions, or even generate new ones. 

### The easisest way to develop an EDA pipeline is to utilize a series of questions to use as tools to guide your investigation. These questions should help focus your attention on certain aspects of your data, and ultimately help you decide what types of tables, graphs, and statistical models to make. 

#### Generating summaries of your data

#### Data visualization:

  -   Summarize tabular data
  -   Visualize variation of single variables
  -   Visualize covariation of multiple variables
  -   Lots of coding


#### 1:  install the following packages
  - dplyr
  - tidyr
  - ggplot2
  - gridExtra
  - GGally

##### Following the example below, install the remaining packages
```{r}
#example
#install.packages('dplyr')

```


##### Solution 1:
```{r}
# install.packages(dplyr) # data manipulation
# install.packages(tidyr) # data manipulation
# install.packages(ggplot2) #plotting
# install.packages(gridExtra) #plotting
# install.packages(GGally)
```

#### 2. Load the packages with the ```library()``` function. Please also load the library ```readr```. Note, unlike the ```install.packages()``` command, you do not have to put package names in quotes or apostrophes when using ```library()```.

##### Response 2:
```{r}
#example
library(readr)
```

##### Solution 2:
```{r}
library(readr) #used to load csv files
library(dplyr) # data manipulation
library(tidyr) # data manipulation
library(ggplot2) #plotting
library(tinytex)
#library(gridExtra) #plotting
#library(GGally) #plotting
```

#### 3: Load dataset. Use the ```read_csv()``` to load the following dataset: hearingLoss.csv and name the dataset ```df```

##### Response 3
```{r}
#Load  data set here. 

```

##### Solution 3:
```{r}
df = read.csv('hearingLoss.csv')
```

### the below list contains data manipulation and helper functions to insure all excercises work properly below. Please make sure to run this. You do not need to understand the code at this time.
```{r}
df = df %>%
  mutate(id = as.factor(id),
         sex = as.factor(sex),
         HearingAidUse = as.factor(HearingAidUse),
         hl_mild = as.factor(hl_mild),
         hl_mod = as.factor(hl_mod),
         NIH_mod = as.factor(NIH_mod),
         NIH_severe = as.factor(NIH_severe),
         sr_hl = as.factor(sr_hl),
         sr_hl2 = as.factor(sr_hl2)) %>%
  select(-X)

df_long = df %>%
  select(id, age, sex,  lpta, rpta) %>%
  gather(ear, pta, lpta:rpta) %>%
  mutate(ear = ifelse(ear == "lpta", "Left Ear", "Right Ear"))
```

#### 4: After loading the dataset we can use a series of commands to examine the structure of the data, the names of the variables within the dataframe, and provide summary statistics of the data.

##### 4.1 Use the ```str()``` command to look at the shape of df, learn the variable names, and variable types. Then, answer the questions below.

##### Response 4.1
```{r}

```
##### Solution  4.1
```{r}
str(df)
```

##### 4.1.1: How many variables are in the dataset?
> Response:

##### 4.1.2: How many observations are there for each variable? 
>

##### 4.1.1-Solution: How many variables are in the dataset?
> Response: The first line that is output after running ```str(df)``` can be used to answer this question along with 4.1.2. there are 17 variables.

##### 4.1.2-Solution: How many observations are there for each variable? 
> 111 observations
#### Data Visualization:

#### The ```head()``` command is also a useful tool for gaining an understanding of the structure of your data. By default it outputs the first 6 observations from each variable in your dataset.

##### 4.2: Use the ```head()``` to examine the first 6 rows of your data. 

```{r}


```

##### 4.2 solution

```{r}
head(df)
```

#### Next we can investigate the variation of each variable. The ```summary()``` function is extremly useful for this. You can run the summary function  on a single variable within a dataset or all variables within the dataset. We will do both.


##### 5.1: use ```summary()``` to produce summary statistics for all variables for df. This is done by placing the dataframe name within the function itself.

```{r}

```


##### 5.1 solution
```{r}
summary(df)
```

##### 5.1.1 Using the output from summary, identify and report the mean age of the participants in the study?

##### 5.1.1 solution:
the mean age of the participants in this study is 74.15.


##### 5.2: Generate a summary of left pure tone average (variable name = lpta).
```{r}
summary(df$lpta)
```

#### 6: Summarizing categorical variables. As can be seen the summary function does not provide very useful information regarding categorical variables such as ```id```. It is worth noting that it is common to have multiple participant ids in the same column if there is an additional variable in the dataframe denoting timepoint. Use the the combination of ```length()``` and ```unique()``` to identify the number of participants in the study.

**Hint, you will need to nest one function inside the other.

```{r}

```

##### Solution 6: 
```{r}
length(unique(df$id))
```

## Onto visualization.

### Univariate visualizations: 
> There are many ways to visualize data in R. Including the base R functions and ggplot2 (gg = grammer of graphics). In this workshop we will rely heavily on the ggplot2 package because of its ability to generate high resolution images that are publication ready. 

#### Ogres are like onions, they come in layers.  ggplots are like ogres, they also come in layers. 

> When generating plots within the ggplot framework it will be helpful to remember that each time you use the + sign, you are adding a new layer to the figure. 

> The first line of code when creating a figure in ggplot2 is ```ggplot()```. The first argument that must me met within this function is telling ggplot where the data is coming from. 

##### 7: Fill in in the blank of the function below and run the command to generate the ggplot template for most figures. This will look like a white piece of paper, a perfect starting point for our figures.

```{r}
#ggplot(data = )
```

##### Solution 7: 
```{r}
ggplot(data = df)
```

#### ```ggplot(data, aes())```
> The second argument within the ```ggplot()``` command is ```aes()```. aes stands for aesthetics and is a function itself.  Within the aes section you can define you x and y variables.
  - ```aes()``` follows a ```,``` after the data's name
  - ```aes()``` is nested within both of the partheses corresponding to the ```ggplot()``` function. 
  - variable names within aes do not require the dataframes name followed by the dollar sign because you have already told ggplot the name of the dataset.
  - declare x and y variables within the aesthetic function like this: aes(x = NameOfVariable, y = NameOfVariable)
  
##### 8: Use ```aes()``` within the ```ggplot()``` command to declare ```age``` as the x axis variable.  

```{r}

```

##### Solution 8:
```{r}
ggplot(data = df, aes(x = age))
```

##### 8.1: What do you see? What did the additional layer add to the figure?

> After declaring your data and the variables of interest within ggplot, as done above and provided in solution 8, we can tell ggplot what type of plot we want to produce. Nearly always, if not always, this command will start with ```geom_``` and following the underscore will by the type plot you want to make.
  - these commands follow the root ggplot command above following a ```+```` sign. It is good practice to hit return after the plus sign.

##### 9: 
> The histogram. A histogram is a popular and useful figure for identifying the type of distribution your variables are. the command. the command to generate a histogram is      
```geom_histogram()```. You can also use the function ```geom_bar()```

  - You do not need to write the variable name within geom_point() because you have already told ggplot both the data's name and the variable of interest. 
  
##### Add the geom_histogram function to the above code to generate a histogram for age. 

##### Response 9.1
```{r }

```

##### Solution 9.1
```{r histogram }
ggplot(data = df, aes(x = age)) + 
  geom_histogram()
```
##### 9.2: Generate the same plot with the ```geom_bar``` function instead of ```geom_histogram```

##### Response 9.2
```{r histogram }

```

##### Solution 9.2
```{r histogram }
ggplot(data = df, aes(x = age)) + 
  geom_histogram()
```

##### 10. Customizing plots. Ggplots are highly customizable. You can change nearly every aspect of any figure. 

##### 10.1: Let's start by changing the color of the border lines of the histogram. To compelte this task, add ```color = ``` within the ```geom_histogram()``` line of your plot.  Following the equal sign of color, type the name of a color of your choice within quotation marks. 
  -Some colors are not available by namee. If you get an error message here read the message and decide if it is because the color was not recognized.
  
##### Response 10:
```{r}

```

##### Solution 10:
```{r}
ggplot(data = df, aes(x = age)) + 
  geom_histogram(color = "blue")
```

##### 10.2: Change the fill color your histogram by adding ```fill = ``` after the color command with a color of your chosing. 
  - Don't forget a comma! 
  - Don't forget quotes
  
##### Response 10:
```{r}

```

##### Solution 10:
```{r}
ggplot(data = df, aes(x = age)) + 
  geom_histogram(color = "blue", fill = "light blue")
```

##### Adding labels.
> The following labels can be added to your ggplot
  - ```title = ``` - Name of the figure, displayed on top
  - ```subtitle = ``` - displayed below the title
  - axis labels - customize x and y axis labels with ```x = ``` and ```y = ```
  - ```caption = ``` - Text will be displayed in the bottom right of the plot
  - ```tag = ``` - text will be displayed in the top left of the plot
  
> Labels are added with the ```labs()``` command. 
  - ```labs()``` should follow the ending parenthese of your geom function. In the working example, following ```geom_histogram```
  - The above commands all go within the labs function.
  - All text for each label type should be within quotation marks.
  - Spaces are a-okay in labels! 
  - Following each label name and the ending quotation mark add a comma and hit return to add your next variable. 
  
  
##### 11.1: Add a title to your plot, be short and descriptive, for example "Histogram of age"

##### Response 11.1
```{r}

```

##### Solution 11.1
```{r}
ggplot(data = df, aes(x = age)) + 
  geom_histogram(color = "blue", fill = "light blue") +
  labs(title = "Histogram of age") 

```

##### 11.2: Change the x and y axis labels to ```Age in years``` and ```Count```. Do not remove the title.
- Why is it important to change the x axis variable name in this case?

##### Response 11.2
```{r}

```

##### Solution 11.2
```{r}
ggplot(data = df, aes(x = age)) + 
  geom_histogram(color = "blue", fill = "light blue") +
  labs(title = "Histogram of age",
       x = "Age in years",
       y = "Count")
```


##### Bringing it all together.   Independent excercise 1:
> Use ggplot to create a histogram for left pure tone average. This variable is labeled lpta. 
  - Modify the color of the fill and color of the bins
  - Generate a meaningful title
  - Change x axis to "Left ear pure tone average"
  - Change y axis to "Number of Observations"
  - Bonus 1:  within geom_histogram add the ```bins = ``` function to change the number of bins to 50. 
      - Hint: no quotes here.
  - Bonus 2: add a density curve over your histogram with geom_dens
      - hint: You will need to google how to complete this step.
      - hint 2: remember to include keywords in your search cush as "ggplot" "histogram" and "density curve". 

##### Independent excercise solution 1
```{r}

```
  
##### Independent excercise solution 1:
```{r}
ggplot(data = df, aes(x = lpta)) + 
  geom_histogram(color = "red", fill = "blue", bins = 50, aes(y = ..density..)) +
  geom_density() +
  labs(title = "Histogram of left pure tone average",
       x = "Left ear pure tone average",
       y = "Number of Observations")

```


#### Plotting two variables:

##### Independent excercise 2: What is the code below doing? Use your own words.  What has it achieved and not achieved?
  -Hint: compare this histogram to the above histogram of lpta.
```{r}
ggplot(data = df) +
geom_histogram(aes(x= lpta)) +
geom_histogram(aes(x = rpta))
```
##### Independent excercise 2.2: Try adding colors to see if that helps.


##### Response independent excercise 2.2: 
```{r}

```

##### Solution independent excercise 2.2: 
```{r}
ggplot(data = df) +
geom_histogram(aes(x= lpta)) +
geom_histogram(aes(x = rpta), fill = "blue")
```

##### Independent excercise 2.3: Have any improvements been made as to the interpretability of the plot? How can we further improve the plot?

> Response: 

#####  ```alpha = ``` 
  - is a method utilized to set the opacity of your plotted variables and requires a numberical value between 0 and 1.
  - ```alpha = 1``` is completely opaque while ```alpha = 0``` is translucent. 
  - declare alpha inside your geom_ function, but outside ```aes()```
  
##### Independent excercise 2.4: Add alpha equals 0.5 to the rpta geom.
  - Hint: don't forget the comma

##### Response independent excercise 2.3: 
```{r}

```

##### Solution independent excercise 2.1: 
```{r}
ggplot(data = df) +
geom_histogram(aes(x= lpta)) +
geom_histogram(aes(x = rpta), fill = "blue", alpha = 0.5)

```
##### Interpretation: Any better? What did we learn? What else does the plot need?
> Response:


##### Adding a legend to wide form data, like we are looking at, is not very easy in ggplot2. You'll cover wide vs long form data tomorrow.
> Your dataframe structure is very important when you plot multiple variables at once. As seen above, we do not want to just stack two histograms together, in this case we lose a lot of information. We need to add both color and changes in opacity. In the following excercise we will use a new dataframe titled df_long. Prior to plotting, check out the dataset. It has less variables than df to help focus your attention on a few key variables. Use your summary commands to report the differences. Pay specific attention to variables ```ear``` and ```pta```. 
- trying using the ```unique()``` command on ear as well. 

##### 12: Investigate the structure df_long
> you will learn how to summarize variables in long form in a later lecture. 

##### Response 12
```{r}
#structure of df_long

```

```{r}
#head of df_long

```

```{r}
#uinique obervations in variable ear

```

##### Solution 12
```{r}
#structure of df_long
str(df_long)
```

```{r}
#head of df_long
head(df_long)
```

```{r}
#uinique obervations in variable ear
unique(df_long$ear)
```

##### 13: Revisiting indepdent question 2: Plotting two histograms together.
> With our newly formatted dataset, coding the two histogram figure is not  much different than the single histogram figure. Necessary step:
 - Tell ggplot the variable that dilineates between left and right ear for the pure tone averages.
 - To do this, add the command ```color = ``` within the original aes call of your ggplot.
 - be careful when using this function because the default is to stack the histogram, which results in higher peaks. 
 
 
```{r}
ggplot(data = df_long, aes(x = pta, color = ear)) +
  geom_histogram() +
  labs(title = "Stacked histograms resulting in confusingly high peaks")
```
##### To navigate this stacking issue, add the following into `geom_histogram(position = "identity")`.


##### Response 13.2:
```{r}

```
##### Solution 13.2:


##### 13.2: Chanage the color function to fill. What's the difference? 

##### Response 13.2:
```{r}

```
##### Solution 13.2:
```{r}
ggplot(data = df_long, aes(x = pta, color = ear)) +
  geom_histogram(position = "identity")
```
##### Note that the figure is still not very interpretable. As above, we are not left with the two histograms on top of eachother. Two steps to fix this as above, add color and change alpha.  In the excercise below:
  - Change the fill color of the plot
  - Change the alpha level to ~ 0.5


  
#### 13.3 Try an alpha level of 0.5 for the above plot utilizing fill, not color.

##### Response 13.3:
```{r}

```

##### Solution 13.2:
```{r}
ggplot(data = df_long, aes(x = pta, fill = ear)) +
  geom_histogram(position = "identity", alpha = 0.5)
```


##### 14: Overlapping figures with alpha in black and white coloring can create figures that are hard to interpret. 

##### *Thought exercise* Why should we care about whether a figure that looks presentable in color is presentable in black and white?

##### One way to get around these issues is to create a figure with multiple plots imbedded in it. Adding `facet_wrap()` following your ggplot with some slight modifications to your figures syntax is a great way to achieve this task. 
  - Like `color()` and `fill()` , `facet_wrap()` creates plots based on a categorical variable such as `ear`.  
  - The syntax is slightly different in this case and requires a `~` followed by the grouping variables name. 
    - e.g. `facet_wrap(~ear)`
    
    
##### Add facetwrap to the above plot and remove all functions adding color and modifying opacacity in the figure The end result should be two histograms in black and white. In addition use the `labs()` command to add a title and change the axis labels. Minimally, change `pta` on the x axis to "Pure tone average" or, for an extra challenge change "pta" to "pta*" and add a subscript to the x axis plot to write out the long form of the acronym. 
  -hint: Don't forget we are always adding layers to your original blank canvas generated by `ggplot(data=df_long, aes(x = pta))`. facet_grid is just another layer. 


##### Response 15:
```{r}

```

##### Solution 15:
```{r}
ggplot(data = df_long, aes(x = pta)) +
  geom_histogram() +
  facet_grid(~ear) +
      labs(title = "Pure tone averages of study participants",
       x = "Pure tone average",
       y = "Count")

```
### Multivariate plots:
> There are many ways to plot multivariate comparisons of variables. The next section will focus on scatter plots, using the `geom_point()` function. Instead of just providing an x axis variable in the main ggplot function we will now have to add a y variable ``ggplot(data = df, aes(x = var1, y = var2)).

##### Excercise 15: Generate scatterplot with ggplot
> Use dataframe `df` to generate a scatterplot comparing age and pure tone average (pta). Don't forget to add axis labels and a title. 

###### Response 15:
```{r}

```

###### Solution 15
```{r}
ggplot(data = df, aes(x = pta, y = age)) +
  geom_point() +
  labs(title = "Scatterplot: L/R puretone average and age",
       x = "Pure tone average",
       y= "Age in years")
```

##### We can now add another dimension to the plot by changing the color of the plot points or the shape of the points. 

##### Excercise 16: 
> Following the examples above, change the fill color of the plot points to represent the sex of the participant. 
###### Response 16:
```{r}

```

###### Solution 16
```{r}
ggplot(data = df, aes(x = pta, y = age, color = sex)) +
  geom_point() +
  labs(title = "Scatterplot: L/R puretone average and age",
       x = "Pure tone average",
       y= "Age in years")
```

##### Excercise 16: 
> To avoid problems of creating plots in color, change the `color` command to `shape` 

###### Response 16:
```{r}

```

###### Solution 16
```{r}
ggplot(data = df, aes(x = pta, y = age, shape = sex)) +
  geom_point() +
  labs(title = "Scatterplot: L/R puretone average and age",
       x = "Pure tone average",
       y= "Age in years")
```

#### We can add another dimension to the plot by faceting the scatterplot based on another categorical variable.  
> Following examples above, add ```facet()``` wrap

##### Excercise 17: Create a faceted version the above plot with the "HearingAidUse" variable. 

###### Response 17:
```{r}

```

###### Solution 17

```{r}
ggplot(data = df, aes(x = pta, y = age, shape = sex)) +
  geom_point() +
  labs(title = "Scatterplot: L/R puretone average and age",
       x = "Pure tone average",
       y= "Age in years") +
  facet_grid(~HearingAidUse)
```
