---
title: "ggplotlecture"
author: "AMS"
date: '2022-07-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data visualization:

-   The practice of translating information into a visual context

-   Helps make complex data easier to understand

### Goals:

-   Teach you how to produce useful and meaningful graphics in R

-   Understand when and when not to use a graphic.

-   Lots of coding on the participant end.

### Libraries

The required libraries for this tutorial include `ggplot2` . As learned in previous lectures, please install and load the library below.

Good practice when working in rmarkdowns is to `#comment` out any installation lines of code, because some installs can be lengthy.

```{r}
#install ggplot2 below.

#load the library ggplot2 below
```

```{r}
#install ggplot2 below.
#install.packages('ggplot2')
#load the library ggplot2 below
library(ggplot2)
library(tidyverse)
```

### Load data

To provide some new code, today we will load data directly from github. Instead of providing a path on your local machine, we will just provide the website and the type of datafile.

```{r}
meir_osf = knitr::include_url('https://osf.io/3eg4x/wiki/home/')
```

#### About the data 

The data to be loaded below comes from 34 individuals with aphasia.

```{r}
df = read.csv("https://raw.githubusercontent.com/rbcavanaugh/pitt-csd-r-workshop/master/DataViz/osf_meir%20and%20Bangash.csv")
```

### Take a peak at the data

```{r}
glimpse(df)
```
