---
title: "Putting it all together"
author: "Rob Cavanaugh"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

The goal of this exercise is to combine skills in data wrangling (part 1) and visualization (part 2) to put together a fully reproducible document that can be shared with collaborators. 

This exercise will just include the basics, but there is a lot more to Rmarkdown. 

See here: https://bookdown.org/yihui/rmarkdown-cookbook/ and here: http://rmarkdown.rstudio.com

## About of R Markdown

### Basic features

There are a couple of features of R-markdown that are important to know

- objects are encapsulated to the document - it won't use any objects or functions already loaded into your workspace.
- content inside the R "chunks" will be run as scriped R.
- content outside the R "chunks" will be compiled as markdown.
- R code can be run "inline" using ESCAPED INLINE CODE GOES HERE

### Code chunk options

Each code chunk has a number of options related to how to display the code, results, warnings, and messages
- WARNING, MESSAGE, ECHO, EVAL, INCLUDE etc. and fig.width out.width etc. 

### YAML

The \-\-\- section above is called the "YAML" header, and contains information about the document
and information that helps to compile the document. This is where you can include the title, author list, and date. 
It's also where you should specify the type of document you'd like to create: an html_document, word_document, or pdf_document. 
Note that for PDF documents, you need to have latex installed (https://yihui.org/tinytex/). 

### Markdown basics

 -  \#, \##, \###, & \#### are headers. so "# Purpose" will place "Purpose" as a level 1 header in the document
 -  \* \**, and \*** are bold, italic, and bold-italic
 - \> will create an indented block quote
 - stuff for superscript, subscript etc. 
 
### More advanced markdown

 - markdown can also incorporate citations, math notation/formulas, and raw latex code

### Other stuff 

- You can run chunks without running the entire document

- Hit knit to run the entire document. 

- Rmarkdown has a visual editor that is slightly more like word


## Goals of this exercise

- Create an Rmarkdown document that can be reproducibly shared with collaborators
- Read in and clean a dataset of your choice (This example includes a dataset from talkbank.org)
- Create at least 2 plots visualizing important aspects of your data
- create a table summarizing your dataset using Kable or paged_table
- Use inline r code to write an automated summary of your dataset (using descriptive statistics)
- Include 1 citation in your document (also....zotero?)...not sure I want to add this. might be too much
- document aspects of your software at the end of the document'
- annotating data - commenting/describing steps
- table of contents/ minor modifications to the YAML
- note 'best practices' throughout the document

# Example code/product

## Loading packages

```{r load packages, warning = FALSE, message = FALSE}
library(tidyverse)
library(here)
```

## Loading data

```{r}
demographics = read.csv(here("rmarkdown", "drouin-2021", "TIMETIME-Demographics.csv"), stringsAsFactors = TRUE)
ganong = read.csv(here("rmarkdown", "drouin-2021", "TIMETIME-Ganong.csv"), stringsAsFactors = TRUE)
pr = read.csv(here("rmarkdown", "drouin-2021", "TIMETIME-PR.csv"), stringsAsFactors = TRUE)
ltrs = read.csv(here("rmarkdown", "drouin-2021", "TIMETIME-LTRS.csv"), stringsAsFactors = TRUE)

```



# Do stuff with Data

# A few plots

# inline text description