---
title: "Exercise_ANSWERKEY_Data Transformation, Tidying, and Wrangling"
author: "Emily Goldberg"
date: '2022-07-25'
output:
  rmdformats::html_clean:
    toc: true
    toc_depth: 2
    code_folding: show
---

Questions - should i set up a few areas for failure?? example is renaming ID when using rbind. or typo in headers
decision: be transparent about building in these issues
for google question - what was the text of the search where you found the answer? were any searches that were unhelpful. can share with class. what modifications were helpful to get to your answer? 

after run it, insert text saying "what was the error message that you received? what did you learn from it?" can you ID discrepancy in data

# Data Transformation, Tidying, and Wrangling Excercise

The best way to learn how to use the functions that we just discussed is to use them in practice, with a specific goal in mind. Remember the case study presented to you at the beginning of our module? That case study officially is about YOU! You were given data for 5 research participants and your mentor asked you to tidy the data for next week's meeting so that you can analyze them together. 

Your mentor says they would like your final product of the aggregated data to be in long format, free of spelling/character errors, and without 2nd rater scores included. 


## About the data

Partnered with a local hospital in the town where you just started your Ph.D. program, your mentor just finished a 5 year clinical trial that investigated the efficacy of a language intervention in people with chronic post-stroke aphasia. The treatment is called Semantic Feature Analysis (SFA), and was designed primarily to improve single word noun naming in individuals with aphasia. Given that SFA is aimed at improving word retrieval, subjects were tested on their naming abilities of treatment list targets before treatment (three times), immediately after treatment (three times), and at a 2 month follow up visit (1 time). 

Every subject was treated on the same words. The duration of intervention lasted 4 weeks. Over these four weeks, participants were treated on either 2 or 3 word lists. Whether or not participants were treated on 2 vs. 3 lists depended on how successful they were in advancing through / responding to intervention. Each treatment list contains 10 items; 5 of the items were directly treated during intervention (T), and 5 were untreated (U) to gauge generalizability. The probes given multiple times before treatment, multiple times immediately after treatment, and once at a follow up time point contained a total of 30 words.

Your mentor is interested in examining changes in naming accuracy over time when comparing the time points mentioned above (pre-treatment, right after treatment, followup). For each subject, probes 1, 2, and 3 were given before treatment; probes 4, 5, and 6 after; and probe 7 at follow up. Your mentor is also interested in associating pre-treatment measures of aphasia severity and other non-language cognitive functions with naming performance before and after treatment. 

In the "data" folder that your mentor sent you, each subject has their own folder. Within the subject folders you will find a .csv file containing the probe naming data. Additionally there is a .csv file containing raw aphasia severity, verbal memory and visuospatial memory, and sustained attention scores titled "cognitive_data.csv". 

Below is a list of the variables included in the 5 "subject##_probes" .csv files: 

* Probe: (the word being probed)
* List: (the treatment list the word comes from)
* Type: (whether the word is is treated T or untreated U)
* Rater1_## and Rater2_##: (two team members alternated scoring performance - Rater 1 and Rater 2 are two different people)
* _##: (the number following "Rater#_" represents the probe number in order)


Your mentor helped you by emailing you a list of variables that the tidy version of this data should contain:

* id: (subject ID)
* item: (the word being probed)
* probe_number: (e.g. 01, 02, 03, 04, 05)
* study_phase: (if before treatment, "pre"; if right after treatment, "post"; if at followup, "followup")
* type: (treated or untreated)
* accuracy (0 or 1)
* aphasia_severity
* verbal_mem
* visual_mem

They specifically asked if the column for "id" could be the VERY FIRST column in the data frame. Also, they asked for categorical variables to be coded as factors. 

To best assist you, I have laid out a list of steps to take in order to wrangle the data. These steps will guide you through the process and assist you with selecting which commands to use. **HOWEVER** I have left out just a few intermediate steps that must be taken to successfully wrangle the data. If you are attempting to complete a step, and receive an error message, I challenge you to (1) read what the error message is telling you and (2) consider what intermediate step might need to be taken to eliminate errors. 

After you submit your final product, the module instructor is going to review your work; students with perfect products will receive a prize!

You better start reshaping these data....you got side tracked with a brutal statistics assignment and a heavy reading for your neuroscience class, and your meeting with your mentor is in 2 days!


### Load libraries:
```{r}
## This chunk is reserved for loading libraries
library(here)
library(tidyverse)
```


## Extreme Makover: Data Edition

To tackle this issue, we first should create code that successfully transforms one subject's data; hopefully once this is accomplished we can replicate the code with other subjects. 

### SUBJECT 01

STEP 1: Read in the probe data for Subject 1

```{r}
here()
subject1_probes <- read.csv(here("M3_DataWrangling", "exercise", "data", "Subject_1", "Subject1_30item.csv"))
```

STEP 2: Become familiar with the data structure

```{r}
head(subject1_probes)
```

Step 3: Pivot the 10-item probe for Subject 1 from wide to long format; in this process, create a new column for probe number (the numbers are included in wide form column headers) and make sure the column containing words is named "item"

```{r}
subject1_probes %>%
  pivot_longer(
    cols = 4:10,
    names_to = c("rater", "probe_number"),
    names_sep = "_",
    values_to = "accuracy"
  ) -> subject1_probes_long

```

Step 4: Rename the "Probe" column to say "item"

```{r}
subject1_probes_long %>%
  rename(item = "Probe") -> subject1_probes_long

```


Step 5: Add a column for subject ID that comes BEFORE item. To specify where a mutated column goes you can use the `.after` or `.before` arguments - hopefully you googled something to solve this such as "mutate place column in location" or "specify column before R mutate" etc. 

```{r}
subject1_probes_long %>%
  mutate(id = "subject01", .before = "item") -> subject1_probes_long
```


Step 6: Add a column that classifies if a probe was given "pre", "post", or "followup". 

```{r}
## First, when I got to know my data I realized that probe number is a character not numeric! I must change to as.numeric: 

subject1_probes_long$probe_number <- as.numeric(subject1_probes_long$probe_number)

## Now I can add my column:


subject1_probes_long %>%
  mutate(
    study_phase = case_when(
    probe_number == 7 ~ "followup",
    probe_number >= 4 ~ "post",
    probe_number <=3 ~ "pre"
  )
  ) -> subject1_probes_long
```

Step 7: Run a few checks to make sure there are no typing errors or other errors within this participant's data:

We know there were 7 probes, and 30 total items. 
```{r}
7*30
```

Should be 210 observations, and there are. 

Let's set item to be a factor and make sure there are not any spelling errors there; if there are 3 lists with 10 items each, there should only be 30 levels for this factor: 

```{r}
subject1_probes_long$item <- as.factor(subject1_probes_long$item)
levels(subject1_probes_long$item)
```

There are 30 total levels. 

Let's count to make sure there are 15 * 7 trained and untrained targets: 

```{r}
15*7 ## 105

subject1_probes_long$Type <- as.factor(subject1_probes_long$Type)

subject1_probes_long %>% count(Type)
```

Things look good. Going to remove the messy data frame: 

```{r}
rm(subject1_probes)
```

### SUBJECT 02

Now that subject 1's data has been tidied, you're ready to replicate this process with subjects 2, 3, 4, and 5! Be warned...you should still carefully look at each subjects' raw messy data frame just in case slight differences exist. 

Step 8: Apply same procedure to Subject 02

```{r}
## READ IN CSV
subject2_probes <- read.csv(here("M3_DataWrangling", "exercise", "data", "Subject_2", "Subject2_30item.csv"))

## view data
head(subject2_probes)
```
I notice the "List" variable has two periods after it. Let's rename that: 
```{r}
subject2_probes %>%
  rename(List = "List..") -> subject2_probes
```

Next pivot longer: 
```{r}
subject2_probes %>%
  pivot_longer(
    cols = 4:10,
    names_to = c("rater", "probe_number"),
    names_sep = "_",
    values_to = "accuracy"
  ) -> subject2_probes_long

head(subject2_probes_long)
```
Rename probe column: 
```{r}
subject2_probes_long %>%
  rename(item = "Probe") -> subject2_probes_long
```

Add column for subject id

```{r}
subject2_probes_long %>%
  mutate(id = "subject02", .before = "item") -> subject2_probes_long
```

Add probe timepoint classification: 

```{r}
## First, change probe to as.numeric: 

subject2_probes_long$probe_number <- as.numeric(subject2_probes_long$probe_number)

## Now I can add my column:

subject2_probes_long %>%
  mutate(
    study_phase = case_when(
    probe_number == 7 ~ "followup",
    probe_number >= 4 ~ "post",
    probe_number <=3 ~ "pre"
  )
  ) -> subject2_probes_long


```

Run checks - there are 210 observations of 8 variables which is a good start
```{r}
subject2_probes_long$item <- as.factor(subject2_probes_long$item)
levels(subject2_probes_long$item)
## 30 levels for item

subject2_probes_long$Type <- as.factor(subject2_probes_long$Type)
subject2_probes_long %>% count(Type)
## 150 trained 150 untrained!
```

```{r}
rm(subject2_probes)
```


### SUBJECT 03

Step 9: Apply same procedure to Subject 03

```{r}
## READ IN CSV
subject3_probes <- read.csv(here("M3_DataWrangling", "exercise", "data", "Subject_3", "Subject3_30item.csv"))

## view data
head(subject3_probes)
```
Same issue with list.

```{r}
## Rename list variable
subject3_probes %>%
  rename(List = "List..") -> subject3_probes

## Pivot longer
subject3_probes %>%
  pivot_longer(
    cols = 4:10,
    names_to = c("rater", "probe_number"),
    names_sep = "_",
    values_to = "accuracy"
  ) -> subject3_probes_long

## Rename probe to item
subject3_probes_long %>%
  rename(item = "Probe") -> subject3_probes_long

## Add id column
subject3_probes_long %>%
  mutate(id = "subject03", .before = "item") -> subject3_probes_long

## Add probe timepoint classification
subject3_probes_long$probe_number <- as.numeric(subject3_probes_long$probe_number)

subject3_probes_long %>%
  mutate(
    study_phase = case_when(
    probe_number == 7 ~ "followup",
    probe_number >= 4 ~ "post",
    probe_number <=3 ~ "pre"
  )
  ) -> subject3_probes_long

```

Run some checks. 210 observations of 8 variables.

```{r}
subject3_probes_long$item <- as.factor(subject3_probes_long$item)
levels(subject3_probes_long$item)
## 30 levels for item

subject3_probes_long$Type <- as.factor(subject3_probes_long$Type)
subject3_probes_long %>% count(Type)
## 105 trained 105 untrained!

```
```{r}
rm(subject3_probes)
```

### SUBJECT 04

Step 10: Apply same procedure to Subject 04


```{r}
## READ IN CSV
subject4_probes <- read.csv(here("M3_DataWrangling", "exercise", "data", "Subject_4", "Subject4_30item.csv"))

## view data
head(subject4_probes)
```


```{r}

## Pivot longer
subject4_probes %>%
  pivot_longer(
    cols = 4:10,
    names_to = c("rater", "probe_number"),
    names_sep = "_",
    values_to = "accuracy"
  ) -> subject4_probes_long

## Rename probe to item
subject4_probes_long %>%
  rename(item = "Probe") -> subject4_probes_long

## Add id column
subject4_probes_long %>%
  mutate(id = "subject04", .before = "item") -> subject4_probes_long

## Add probe timepoint classification
subject4_probes_long$probe_number <- as.numeric(subject4_probes_long$probe_number)

subject4_probes_long %>%
  mutate(
    study_phase = case_when(
    probe_number == 7 ~ "followup",
    probe_number >= 4 ~ "post",
    probe_number <=3 ~ "pre"
  )
  ) -> subject4_probes_long

```

Run some checks. 210 observations of 8 variables.

```{r}
subject4_probes_long$item <- as.factor(subject4_probes_long$item)
levels(subject4_probes_long$item)
## 30 levels for item

subject4_probes_long$Type <- as.factor(subject4_probes_long$Type)
subject4_probes_long %>% count(Type)
## 105 trained 105 untrained!

```
```{r}
rm(subject4_probes)
```

### SUBJECT 05

Step 11: Apply same procedure to Subject 05

```{r}
## READ IN CSV
subject5_probes <- read.csv(here("M3_DataWrangling", "exercise", "data", "Subject_5", "Subject5_30item.csv"))

## view data
head(subject5_probes)
```

```{r}
## Pivot longer
subject5_probes %>%
  pivot_longer(
    cols = 4:10,
    names_to = c("rater", "probe_number"),
    names_sep = "_",
    values_to = "accuracy"
  ) -> subject5_probes_long

## Rename probe to item
subject5_probes_long %>%
  rename(item = "Probe") -> subject5_probes_long

## Add id column
subject5_probes_long %>%
  mutate(id = "subject05", .before = "item") -> subject5_probes_long

## Add probe timepoint classification
subject5_probes_long$probe_number <- as.numeric(subject5_probes_long$probe_number)

subject5_probes_long %>%
  mutate(
    study_phase = case_when(
    probe_number == 7 ~ "followup",
    probe_number >= 4 ~ "post",
    probe_number <=3 ~ "pre"
  )
  ) -> subject5_probes_long
```

Run some checks. 210 observations of 8 variables.

```{r}
subject5_probes_long$item <- as.factor(subject5_probes_long$item)
levels(subject5_probes_long$item)
## 30 levels for item

subject5_probes_long$Type <- as.factor(subject5_probes_long$Type)
subject5_probes_long %>% count(Type)
## 105 trained 105 untrained!
```
```{r}
rm(subject5_probes)
```

### Combining subject data

Step 12: Combine the subject data so there is an aggregated data frame. 

In our situation, our data frames contain the same factors with the same exact labels as well as the same number of observations. We don't need to join them based on a common column and instead really just want to stack them. Because of this, we will use `r`bind()`. 

```{r}
aggregated_subject <- rbind(subject1_probes_long, subject2_probes_long, subject3_probes_long, subject4_probes_long, subject5_probes_long)

head(aggregated_subject)
```

If 210*5=1050, it makes great sense that we have 1050 rows of 8 variables. 


### Combine aggregated naming data with cognitive data

Now we need to read in the cognitive and language data, select the columns we wish to use, and combine it with the aggregated subject naming data. 

Step 13: Read in cognitive data

```{r}
cognitive_vars <- read.csv(here("M3_DataWrangling", "exercise", "data", "cognitive_variables.csv"))
```

Step 14: Select only the variables your mentor wants to use - aphasia severity, verbal memory, and visual memory

```{r}
cognitive_vars %>%
  select(ID, aphasia_severity, verbal_mem, visuospatial_mem) -> cognitive_vars_select

head(cognitive_vars)
```

Notice that ID is capitalized here but not in the aggregated subject data. These must match for us to successfully join!


STEP 15: Change ID in cognitive variable data frame to be lowercase: 
```{r}
cognitive_vars_select %>%
  rename(id = "ID") -> cognitive_vars_select
```


Step 15: Join using ID as the common factor!
```{r}
naming_master <- left_join(aggregated_subject, cognitive_vars_select, by = "id")
```

Step 16: Let's do a quick check on the data. What we know about it: there should be 5 subjects; there should be 30 words, 15 trained and 15 untrained; there should be 3 lists; there should be 2 raters; there should be 3 study phase levels. Let's make ID, item, type, rater, and study phase factors: 

```{r}
naming_master$id <- as.factor(naming_master$id)
naming_master$item <- as.factor(naming_master$item)
naming_master$Type <- as.factor(naming_master$Type)
naming_master$rater <- as.factor(naming_master$rater)
naming_master$study_phase <- as.factor(naming_master$study_phase)
```

What do you notice is wrong? Item has 32 levels but should only have 30. Let's look at them:

```{r}
levels(naming_master$item)
```

hot-air-balloon vs. hotairballoon
cruise ship vs. cruiseship


Let's count these differences
```{r}
naming_master %>%
  count(item)
```

Based on this, go ahead and correct the errors!

```{r}
naming_master %>%
  mutate(item = gsub("cruise ship", "cruiseship", item)) -> naming_master

naming_master %>%
  mutate(item = gsub("hot-air-balloon", "hotairballoon", item)) -> naming_master

```

Now change item back to factor and see if your changes solved the issue.

```{r}
naming_master$item <- as.factor(naming_master$item)
levels(naming_master$item)
```

Looks like there are 5 levels for id; there are 2 type levels; there are 2 rater levels; there are 3 phase levels. This all checks out! 

Step 16: Write a csv file so your work is saved

```{r}
naming_master %>%
  write.csv(here("M3_DataWrangling", "exercise", "data", "naming_master.csv"))
```


